---
title: "DATA607 Assignment 2"
author: "Alexander Simon"
date: "2024-02-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("RMySQL", repos="http://cran.us.r-project.org")
```

## Introduction

The goals of this assignment are to collect data about movie ratings, store the data in an SQL database, and then import the SQL data into an R data frame.

## Data collection

I asked 10 friends via text message to rate 6 movies on a scale of 1 (awful) to 5 (excellent), or 0 if they had not watched it (subsequently converted to \\N for SQL).

I listed the movies in alphabetical order to minimize perception of an implied ranking.

-   Avatar 2

-   Barbie

-   News of the World

-   Soul

-   The Sound of Metal

-   Top Gun Maverick

Five replied with ratings, 3 did not reply, and 2 replied but said it was too difficult to sum up movies with a single number. So the overall response rate was 50%.

I saved the ratings in a CSV file called `ratings.csv` (available in my GitHub DATA607 repository).

## Loading the data into MySQL (local database)

I used the RMySQL library to interface with MySQL.

```{r library}
library(RMySQL)
```

First, I established a connection to MySQL. I included prompts for the username and password as a workaround to showing them in the code.

```{r db-connect}
connObj <- dbConnect(RMySQL::MySQL(), user=rstudioapi::askForPassword("MySQL username"), password=rstudioapi::askForPassword("Password"), host='localhost')
```

Next, I created a new SQL database.

```{sql new-dataase, connection=connObj}
CREATE DATABASE movies;
```

Then I created an empty table in the database that matched the structure of `ratings.csv`. Name is a text variable and the ratings are integers. The name is the primary key, so it can't be null, but the ratings may be null.

```{sql create-table, connection=connObj}
CREATE TABLE movies.ratings (
  name VARCHAR(100) NOT NULL,
  rating1 INT(1) NULL,
  rating2 INT(1) NULL,
  rating3 INT(1) NULL,
  rating4 INT(1) NULL,
  rating5 INT(1) NULL,
  rating6 INT(1) NULL,
  PRIMARY KEY (name));
```

I tried importing data from ratings.csv into the movies.ratings SQL table. The code below runs but doesn't add anything to the table.

```{sql import-data1, connection=connObj}
LOAD DATA LOCAL INFILE 'ratings.csv' INTO TABLE movies.ratings FIELDS TERMINATED BY ',' LINES TERMINATED BY '\n';
```

However, if I execute the SQL command on the command line, the movies.ratings table is populated.

```{sql import-data2, eval=FALSE}
mysql> LOAD DATA LOCAL INFILE 'ratings.csv' 
    ->             INTO TABLE movies.ratings 
    ->             FIELDS TERMINATED BY ',' 
    ->             LINES TERMINATED BY '\n';
Query OK, 5 rows affected (0.00 sec)
Records: 5  Deleted: 0  Skipped: 0  Warnings: 0

mysql> SELECT * FROM movies.ratings;
+----------+---------+---------+---------+---------+---------+---------+
| name     | rating1 | rating2 | rating3 | rating4 | rating5 | rating6 |
+----------+---------+---------+---------+---------+---------+---------+
| Akin     |    NULL |    NULL |    NULL |    NULL |    NULL |    NULL |
| Hannah   |    NULL |       3 |    NULL |    NULL |    NULL |    NULL |
| Kimberly |    NULL |    NULL |    NULL |       2 |       4 |    NULL |
| Steph    |    NULL |       3 |    NULL |    NULL |    NULL |       3 |
| Tony     |       3 |    NULL |       4 |       3 |       5 |       5 |
+----------+---------+---------+---------+---------+---------+---------+
```

## Importing data from MySQL to R (local database)

I queried the database to select all rows in the ratings table and then assigned them to a new data frame.

```{r local-get}
ratings <- dbGetQuery(connObj, "SELECT * FROM movies.ratings;")
```

I confirmed that the data frame contained the correct data. Missing data are represented by NA.

```{r dataframe1}
ratings
```

Finally, I closed the database connection.

```{r db-disconnect}
dbDisconnect(connObj)
```

## Creating a Google Cloud MySQL instance and importing data to R

I set up a Google Cloud MySQL database instance and connected to it from MySQL Workbench as explained in the [YouTube video](https://www.youtube.com/watch?v=dNqz9hCQozM) described on Blackboard.

I created a movies database with a ratings table in MySQL Workbench using the script below.

```{sql GCP-script, eval=FALSE}
DROP DATABASE IF EXISTS movies;
CREATE DATABASE movies;
USE movies;

CREATE TABLE ratings
(
	first_name VARCHAR(100) NOT NULL,
    rating1 INT(1) NULL,
    rating2 INT(1) NULL,
    rating3 INT(1) NULL,
    rating4 INT(1) NULL,
    rating5 INT(1) NULL,
    rating6 INT(1) NULL,
    PRIMARY KEY (first_name)
);

INSERT INTO ratings (first_name,rating1,rating2,rating3,rating4,rating5,rating6)
		VALUES 
        ('Steph',NULL,3,NULL,NULL,NULL,3),
        ('Kimberly',NULL,NULL,NULL,2,4,NULL),
        ('Hannah',NULL,3,NULL,NULL,NULL,NULL),
        ('Akin',NULL,NULL,NULL,NULL,NULL,NULL),
        ('Tony',3,NULL,4,3,5,5);

```

I confirmed that the table is correctly populated in the database instance on Google Cloud.

```{sql GCP-table, eval=FALSE}

mysql> SELECT * FROM ratings;
+------------+---------+---------+---------+---------+---------+---------+
| first_name | rating1 | rating2 | rating3 | rating4 | rating5 | rating6 |
+------------+---------+---------+---------+---------+---------+---------+
| Akin       |    NULL |    NULL |    NULL |    NULL |    NULL |    NULL |
| Hannah     |    NULL |       3 |    NULL |    NULL |    NULL |    NULL |
| Kimberly   |    NULL |    NULL |    NULL |       2 |       4 |    NULL |
| Steph      |    NULL |       3 |    NULL |    NULL |    NULL |       3 |
| Tony       |       3 |    NULL |       4 |       3 |       5 |       5 |
+------------+---------+---------+---------+---------+---------+---------+
```

To import the data into R, I established a new database connection to this instance using the public IP address that was provided during setup.

```{r cloud-db-connect}
connObj2 <- dbConnect(RMySQL::MySQL(), user='root', password=rstudioapi::askForPassword("Password"), host='35.202.52.132')
```

The rest of the process to import the data into R is the same as importing from the local database described in the previous section.

```{r GCP-get}
ratings2 <- dbGetQuery(connObj2, "SELECT * FROM movies.ratings;")
```

I confirmed that the data frame contained the correct data.

```{r dataframe2}
ratings2
```

Finally, I closed the database connection.

```{r db-disconnect}
dbDisconnect(connObj2)
```

## Conclusions

I successfully loaded data from a local CSV file into a MySQL database and then imported the data from MySQL into an R data frame. I also successfully created a database in a MySQL instance on Google Cloud and imported data from it into an R data frame.
