---
title: "tidyverse CREATE: using the `purrr` package with data"
output: html_document
date: "2024-04-28"
author:
  - name: "AJ Strauman-Scott"
  - name: "Extended by Alexander Simon"  
    affiliations:
      - name: City University of New York SPS
        department: Department of Data Analytics
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

This paper will use a sample dataset from Kaggle.com to create a programming sample “vignette” that demonstrates how to use the map functions of the `purrr` TidyVerse package [@purrr].

The 'purrr' package loads automatically as part of the `tidyverse` series of R packages.

```{r load-libraries, message=FALSE}
library(tidyverse)
```

We load a dataset from Kaggle.com with data from All data is from American Community Survey 2010-2012 Public Use Microdata Series [@kaggle].

```{r}
majors <- read_csv('https://github.com/annejscott/DATA607/raw/main/assignments/week9/all-ages.csv', col_names = TRUE, show_col_types = FALSE) |> 
  as.data.frame()
```

To create a new column with the percent of graduates from each major who are employed full time, we define a function to calculate the percentage employed, and then use the `purrr` package to `map` that function onto our data (in this case, `purrr::map2_dbl()` specifically.

```{r function-def}
calculate_proportion_employed <- function(full_time_employed, total) {
  proportion <- full_time_employed / total
  return(proportion)
}

majors <- majors |> 
  mutate(proportion_employed = purrr::map2_dbl(Employed_full_time_year_round, Total, calculate_proportion_employed)) |> 
  arrange(desc(proportion_employed))

head(majors)
```

'Management Information Systems and Statistics' and 'Computer and Information Systems' are the majors that have the highest rates of post-graduate employment.

All of our majors are in full capital letters. Let's use `purrr::map_chr()` with `str_to_title()` from the `tidyverse` `stringr` package to change that into proper grammea.

```{r}
majors$Major <- purrr::map_chr(majors$Major, ~str_to_title(.x))

head(majors)
```

We load another .csv file from the same dataset, this time the information on gender breakdown in each major.

```{r}
women_stem <- read_csv('https://github.com/annejscott/DATA607/raw/main/assignments/week9/women-stem.csv', col_names = TRUE, show_col_types = FALSE) |> 
  as.data.frame()
```

The `women_stem` dataset has information on gender proportions that doesn't exist in the `majors` dataset. Even though the datasets have different numbers of observations,the `purrr::map_dbl()` function can map the values from one data set into another.

```{r}
# apply str_to_title to women_stem df so strings in Major column match
women_stem$Major <- purrr::map_chr(women_stem$Major, ~str_to_title(.x))

sharewomen_map <- setNames(women_stem$ShareWomen, women_stem$Major)

majors$ShareWomen <- purrr::map_dbl(majors$Major, ~ sharewomen_map[.x])

head(majors)
```

Using the `purrr` package, we've been able to create an entirely new data set.

<br>

## More about `purrr`

### Using map() to identify missing values

A useful application of `map()` is to determine the number of missing values (NAs) in each column of a dataframe. Because the `majors` and `women_stem` dataframes had different numbers of observations, the application of `map_dbl()` above resulted in some missing values.

The code below shows that there are 97 missing values in the new `ShareWomen` column and no missing values in the other columns.

```{r}
map(majors, ~ sum(is.na(.)))
```

<br>

### Other variants of map()

In addition to `map()` and `map2()`, which take one or two lists as input, respectively, there are many variants including:

-   `pmap()` - arbitrary number of inputs

-   `modify()` - outputs same data type as input

-   `walk()` - perform an action iteratively but doesn't return anything

Examples of these functions are shown below.

<br>

#### pmap()

The code below concatenates all information about each major (code, name, category) in the `women_stem` dataframe into a single string.

```{r}
major_description <- function(Major_code, Major, Major_category, ...) {
  major_str <- paste(Major_code, Major, Major_category, sep = "_")
  return(major_str)
}

women_stem %>% 
  pmap_chr(major_description) %>% 
  head()
```

<br>

#### modify()

The code below uses a conditional form of `modify()` to apply a log transformation to all numeric values in the `women_stem` dataframe. This could be useful to transform data for a plot.

```{r}
# log1p(x) = log(x+1), useful when there are values between 0 and 1 such as proportions
women_stem2 <- modify_if(women_stem, is.numeric, ~ log1p(.x))
head(women_stem2)
```

<br>

#### walk()

The code below divides the `women_stem` dataframe by the major category and saves each part to a separate CSV file using `walk2()`, which takes two lists (categories and file paths) as input. Walk operations can also be performed with one list using `walk()` or many lists using `pwalk()`.

```{r, warning=FALSE}
# Create new folder in working directory
wd <- getwd()
new_folder <- paste0(wd, "/new")
dir.create(new_folder)

# Create a list for each major category
categories <- split(women_stem, women_stem$Major_category)
paths <- file.path(new_folder, paste0("women_stem-", names(categories), ".csv"))
walk2(categories, paths, write.csv)

# List the files
dir(new_folder)
```

<br>

## References

Wickham H. 2019. Functionals. In *Advanced R*, 2nd ed., Chapter 9. Chapman & Hall. <https://adv-r.hadley.nz/functionals.html>

Apply functions with purrr :: Cheatsheet. <https://rstudio.github.io/cheatsheets/html/purrr.html>
